Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-19 E-Hypergraphs-LICS24.tex
minor fixes to completeness proof

To see that the functor is faithful,  we prove that $f \neq g$ in $\textbf{SMT}^+(\Sigma, \mathcal E)$ implies $\llbracket f \rrbracket \neq \llbracket g \rrbracket$.  First,  calculate weak normal form $\Sigma^+$-terms of $f$ and $g$ as $f_1 + \ldots + f_n$ and $g_1 + \ldots + g_m$,  respectively.  Applying the functor $\llbracket - \rrbracket$,  we thus have that
\[
	\llbracket f \rrbracket = \llbracket f_1 \rrbracket + \ldots + \llbracket f_n \rrbracket \qquad
	\llbracket g \rrbracket = \llbracket g_1 \rrbracket + \ldots + \llbracket g_m \rrbracket ~ . 
\]
Observe  that therefore $\llbracket f_1 \rrbracket + \ldots + \llbracket f_n \rrbracket$ and $\llbracket g_1 \rrbracket + \ldots + \llbracket g_m \rrbracket$ are weak normal forms.  We conisder two cases: where $n=m$ and where $n\neq m$.  In the first case,  it is immediate that $\llbracket f \rrbracket \neq \llbracket g \rrbracket $,  since their weak normal forms cannot be equal.  Otherwise,  if $n = m$,  and $\llbracket f \rrbracket = \llbracket g \rrbracket$,  we have that there exists a permutation $\sigma$ such that for every index $i$ we have $\llbracket f_i \rrbracket \neq \llbracket g_{\sigma(i)} \rrbracket $,  and thus by  $(*)$ that $f_i = g_{\sigma(i)}$,  implying that $f = g$ which is a contradiction.

To see that the functor is full,  consider a morphism $f$ in $\WellTypedMdaEcospans$.  We can find an equivalent weak normal form $f_1 + \ldots + f_n$,  where each $f_i$ contains no hierarchical edges.  Thus,  by $(*)$ we can find morphisms $g_i$ in $\textbf{SMT}(\Sigma, \mathcal E)$ such that $\llbracket g_i \rrbracket = f_i$.  Thus,  we can construct a morphism $g_1 + \ldots + g_n$ such that $\llbracket g_1 + \ldots + g_n \rrbracket = \llbracket g_1 \rrbracket + \ldots \llbracket g_n \rrbracket = f_1 + \ldots + f_n$,  as required. 
\end{proof}

>>>>

To see that the functor is faithful,  we prove that $f \neq g$ in $\textbf{SMT}^+(\Sigma, \mathcal E)$ implies $\llbracket f \rrbracket \neq \llbracket g \rrbracket$.  First,  calculate weak normal form $\Sigma^+$-terms of $f$ and $g$ as $f_1 + \ldots + f_n$ and $g_1 + \ldots + g_m$,  respectively.  Applying the functor $\llbracket - \rrbracket$,  we thus have that
\[
	\llbracket f \rrbracket = \llbracket f_1 \rrbracket + \ldots + \llbracket f_n \rrbracket \qquad
	\llbracket g \rrbracket = \llbracket g_1 \rrbracket + \ldots + \llbracket g_m \rrbracket ~ . 
\]
Observe that $\llbracket f_1 \rrbracket + \ldots + \llbracket f_n \rrbracket$ and $\llbracket g_1 \rrbracket + \ldots + \llbracket g_m \rrbracket$ are weak normal forms.  We consider two cases: where $n=m$ and where $n\neq m$.  In the first case,  it is immediate that $\llbracket f \rrbracket \neq \llbracket g \rrbracket $,  since equal terms have weak normal forms with the same number of components.  Otherwise,  if $n = m$,  assume for contradiction that $\llbracket f \rrbracket = \llbracket g \rrbracket$.  We have that there exists a permutation $\sigma$ such that for every index $i$ we have $\llbracket f_i \rrbracket \neq \llbracket g_{\sigma(i)} \rrbracket $,  and thus by  $(*)$ that $f_i = g_{\sigma(i)}$,  implying that $f = g$ which is a contradiction.  Therefore $\llbracket f \rrbracket \neq \llbracket g \rrbracket$ as required.

To see that the functor is full,  consider a morphism $f$ in $\WellTypedMdaEcospans$.  We can find an equivalent weak normal form $f_1 + \ldots + f_n$,  where each $f_i$ contains no hierarchical edges.  By $(*)$ we can find morphisms $g_i$ in $\textbf{SMT}(\Sigma, \mathcal E)$ such that $\llbracket g_i \rrbracket = f_i$.  Thus,  we can construct a morphism $g_1 + \ldots + g_n$ such that $\llbracket g_1 + \ldots + g_n \rrbracket = \llbracket g_1 \rrbracket + \ldots \llbracket g_n \rrbracket = f_1 + \ldots + f_n$,  as required. 
\end{proof}
